---
title: "The Bivariate Binomial Conditionals Distribution (BBCD)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Bivariate Binomial Conditionals Distribution (BBCD)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(BCD)
```

# Introduction

This vignette introduces the Bivariate Binomial Conditionals Distribution (BBCD), defined via conditional specifications, as proposed by Ghosh, Marques, and Chakraborty (2025). The `BCD` package provides functions to evaluate the joint and cumulative distributions, perform random sampling, and estimate parameters via maximum likelihood.

# Joint Probability: `dbinomBCD()`

The joint probability mass function (p.m.f.) of the BBCD is given by:

\[
P(X = x, Y = y) = K \cdot \binom{n_1}{x} \binom{n_2}{y} p_1^x (1 - p_1)^{n_1 - x} p_2^y (1 - p_2)^{n_2 - y} \lambda^{xy},
\]

where \( K \) is a normalizing constant ensuring the probabilities sum to 1.

## Example

```{r}
dbinomBCD(x = 2, y = 1, n1 = 5, n2 = 5, p1 = 0.5, p2 = 0.4, lambda = 0.5)
# independence case
dbinomBCD(x = 2, y = 1, n1 = 5, n2 = 5, p1 = 0.5, p2 = 0.4, lambda = 1.0) 
```

# Cumulative Distribution: `pbinomBCD()`

The function `pbinomBCD()` computes the cumulative distribution:

\[
P(X \leq x, Y \leq y)
\]

## Example

```{r}
pbinomBCD(x = 2, y = 5, n1 = 5, n2 = 5, p1 = 0.5, p2 = 0.4, lambda = 0.5)
pbinomBCD(x = 1, y = 1, n1 = 10, n2 = 10, p1 = 0.3, p2 = 0.6, lambda = 1)
```

# Random Sampling: `rbinomBCD()`

Generate samples from the BBCD using:

```r
rbinomBCD(n, n1, n2, p1, p2, lambda)
```

## Example

```{r}
set.seed(123)
samples <- rbinomBCD(n = 100, n1 = 10, n2 = 10, p1 = 0.5, p2 = 0.4, lambda = 1.2)
head(samples)
```

# Maximum Likelihood Estimation: `MLEbinomBCD()`

Estimate the parameters of the distribution from data.

## Example

```{r}
data <- rbinomBCD(n = 100, n1 = 6, n2 = 4, p1 = 0.6, p2 = 0.3, lambda = 1.5)
fit <- MLEbinomBCD(data)
fit
```

You may also fix known values for `n1` and `n2`:

```{r}
MLEbinomBCD(data, fixed_n1 = 6, fixed_n2 = 4)
```

# Real Data Example 
The dataset `shacc` is related to accident records for 122 experienced railway shunters across two historical periods.
```{r}
data(shacc)
head(shacc)
plot(shacc$X, shacc$Y, xlab = "Accidents 1937–42", ylab = "Accidents 1943–47")
```

```{r}
fit <- MLEbinomBCD(shacc, fixed_n1 = 33, fixed_n2 = 27)
FTtest(shacc, "BBCD", params = fit, num_params = 3)
```
The dataset `seedplant` records the number of seeds sown and the number of resulting plants grown over plots of fixed area (5 square feet).
```{r}
data(seedplant)
head(seedplant)
plot(seedplant$X, seedplant$Y, xlab = "Seeds Sown", ylab = "Plants Grown")
```

```{r}
EstParams <- MLEbinomBCD(shacc, fixed_n1 = 13, fixed_n2 = 11)
FTtest(shacc, "BBCD", params = EstParams, num_params = 3)
```

**Reference:**
Ghosh, I., Marques, F., & Chakraborty, S. (2025). *A form of bivariate binomial conditionals distributions*. Communications in Statistics - Theory and Methods, 54(2), 534–553.
